---
title: "Sprint Qualifying Analysis"
author: "Jose Delatorre"
date: "2022-08-17"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(ggplot2)
library(tibble)
library(dplyr)
library(scales)
```

# Formula 1 Sprint Qualifying Analysis

The purpose of this analysis is to determine how sprint qualifying races for the 2021 and 2022 seasons of Formula 1 (F1) have changed the outcomes of the race weekend and the race itself. The concept of sprint qualifying format for F1 was introduced in the 2021 season changes such that three race weekends in a season will be utilizing the spring qualifying races. 

A normal Formula 1 race weekend consists of about three days of on track time with the cars. Fridays (Day 1) are Free Practice Sessions (FP) where teams are able to test out the track and car setups over two separate 1-hour sessions. Saturdays are a mix of the final FP session for a total of three FP sesssions and three rounds of lap-time based qualifying sessions durin which the bottom 5 are 'eliminated' per session. A drivers elimination from the qualifying session will determine their position on the starting grid for Day 3 which is the Grand Prix running for a minimum of 305km (190miles) plus one lap. The only exception to this metric is the Monaco Grand Prix which is a distance of 260km (160mi).

This race weekend setup is modified slightly for sprint qualifying races. Day 1 is condensed to a single Free Practice session superseded by lap-time based qualifying. Day 2 starts with a second Free Practice session which is followed by the sprint race lasting for a total of 100km (62mi). Day 3 remains unchanged consisting of only the Grand Prix with the starting grid positions now determined by the results of the Sprint Race. For more info regarding Formula 1 Sprint races go here <link for Sprint>

## Ask
### Question 1
How have sprint qualifying weekends for the 2021 and 2022 Formula 1 seasons changed the outcomes of the actual races days?

### Question 2
To be more precise, is the pole sitting driver after the lap-time based qualifying sessions also the driver to win the sprint qualifying race and in turn that respective Grand Prix? 

### Question 3
In the case that the pole position driver changes between the lap-time based qualifying and winner of the sprint race, how many instances does the sprint qualifying winner also win that Grand Prix? 

## Prepare
### Importing the data for analysis
The data that for the analysis was downloaded via [Kaggle](https://www.kaggle.com/datasets/rohanrao/formula-1-world-championship-1950-2020) after which it is loaded in SQL nor and RStudio into a pipeline using read.csv and then previewed with the tibble function. The data is imported using the table function rather than the read.csv as it creates a more mergable table however it is entirely preferential. 

The data for this analysis was downloaded via Kaggle after which it is loaded into SQL and RStudio using a table 

### Dataset 1: Sprint results
```{r}
sprint_results <- data.frame(read.table("F1_Data_1950-2022/sprint_results.csv",header=TRUE, sep=","))

```
### Dataset 2: Formula 1 Race Data from 1950-2022
```{r}
races <- data.frame(read.table("F1_Data_1950-2022/races.csv",header=TRUE,sep=","))

```
### Dataset 3: Comprehensive Driver Data
```{r}
drivers <- data.frame(read.table("F1_Data_1950-2022/drivers.csv",header = TRUE,sep=","))

```
### Dataset 4: Grand Prix Results
```{r}
results <- data.frame(read.table("F1_Data_1950-2022/results.csv", header= TRUE,sep = ","))

```

## Process - Cleaning the Data
After the data has been imported, it is recommended the data be cleaned to remove any unnecessary clutter from the three tables. Cleaning the data will help with reducing the amount of clutter for each of the tables which will become especially important when joining the tables in the next few steps. Taking a look at each of the datasets using the View() function and remove any nulls.

```{r}
sprint_results <- sprint_results[-c(13,16)]
results <- subset(results,select = -c(13,18))
drivers <- subset(drivers, select = -c(dob,url,nationality))
races <- races[-c(8:18)]
```
RUN ONLY ONCE, IT WILL CREATE AN ERROR 

An error may occur when removing columns from the races dataset using column names so to avoid the error and to save space, using their indexed column numbers to remove them is cleaner
### First Join - Sprints Results and Drivers 

Once the unnecessary columns have been removed, join sprint_results and drivers tables based on the driverId, this should then create a much smaller table for us to view the results of the sprint qualifying races with the drivers names shown. 

```{r}
sprints_x_driver <- merge(sprint_results,drivers, by = "driverId")
sprints_x_driver <- merge(sprints_x_driver,races, by = "raceId")
```
Verify that the two tables have merged correctly and then perform a secondary join between 'sprints_merged' and 'races' using "raceId" as the primary key, overwriting the old 'sprints_x_driver' dataset

### Second Join - Drivers and Results

All tables that have been uploaded contain several potential primary keys which could make them suitable for merging into a single table however when attempting to merge them into a single table, the sprint results data merging with the races dataset causes the position to be overwritten. This is an extremely important point of the analysis as the results and winners of laptime qualifying, sprint qualifying, and the Grand Prix may not be the same. While an outer join would also be possible, this leads to 


```{r}
GP_results <- merge(results,drivers, by = "driverId") 
GP_results <- merge(GP_results,races, by = "raceId")
GP_results <- GP_results[-c(3,4,17)]
```

### Confirm join success
Confirm that the second join worked however this time using the tibble function will provide a slightly more detailed

```{r 2021 Sprint Races}

```
### Filter the data
Now that the datasets are condensed into a single working table we can utilize the single table to filter as necessary. The first filter to be applied will be separating the 2021 and 2022 seasons, the second will be by the Grand Prix, the third and final filter will be applied to reduce the analysis to the top five drivers per event.

#### Filter by Grand Prix, Year, and Top Five of Sprint Qualifying

```{r}
Sprint_GP01_2021 <- filter(sprints_x_driver, year=='2021'& circuitId =='9' & position <=5)
Sprint_GP02_2021 <- filter(sprints_x_driver, year=='2021'& circuitId =='14' & position <=5)
Sprint_GP03_2021 <- filter(sprints_x_driver, year =='2021'& circuitId=='18'& position <=5)
Sprint_GP01_2022 <- filter(sprints_x_driver, year=='2022'& circuitId=='21'& position <=5)
Sprint_GP02_2022 <- filter(sprints_x_driver, year=='2022'& circuitId=='70'& position <=5)
```
#### Filter by Grand Prix, Year, and Top Five of Final Race Results

```{r}
Results_GP01_2021 <- filter(GP_results, year=='2021'& circuitId =='9' )
Results_GP02_2021 <- filter(GP_results, year=='2021'& circuitId =='14')
Results_GP03_2021 <- filter(GP_results, year=='2021'& circuitId =='14')
Results_GP01_2022 <- filter(GP_results, year=='2022'& circuitId =='21')
Results_GP02_2022 <- filter(GP_results, year=='2022'& circuitId =='70')
```
#### Top Five Drivers of Each Grand Prix
Grid = Starting grid position
Position = final position at completion of race/sprint
```{r}
RSS <- merge(Results_GP01_2021 ,Sprint_GP01_2021, by = 'driverId')
# Use \\ to bypass "\" which is considered an escape sequence
RSS[RSS == "\\N"] <- 0
transform(RSS, position.x = as.numeric(position.x))
```

## 2021 Sprint Races

For the 2021 Formula 1 Season there were a total of 3 sprint races occurring at rounds 10, 14, and 19 for the British GP, Italian GP, and Sao Paulo GP respectively.

```{r Sprint Races}
ggplot(RSs_GP01_2021, aes(x = code, y = position.x)) +
  geom_point(stat = 'identity', width = .5, fill ='grey')
    facet_wrap(~'grid')
```

## 2022 Sprint Races

At the time of this analysis there has been a total of 2 sprint races occurring at rounds 4 and 11 for the Emilia Romagna GP and Austrian GP respectively. The third and final sprint race of the 2022 season will be held at the Sao Paulo GP.

```{r}

ggplot(RSS,aes(x = code.x, y = position.y)) +
  geom_point(size = 3)+
  geom_segment(aes(x = code.x,
                   xend = code.x,
                   y = position.y,
                   yend = position.x)) +
  labs(title = 'British GP Results',
       caption = 'abc')
  theme(axis.text.x = element_text(angle=65,vjust = 0.7))
```

